Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-09 latex/main.tex

The ARG represents the full evolutionary history of SARS-CoV-2 and
is in close agreement with the state-of-the-art UShER phylogeny built using the Viridian 
dataset\cite{Hunt2024}.
First, we validated the broad structure of the ARG against the UShER phylogeny\cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
Tanglegrams reveal that the ARG captures known phylogenetic relationships
both within sublineages and at the level of major lineages (including VOCs),
with good agreement between the phylogenetic bipartitions in the ARG and the UShER 
phylogeny (Figure~\ref{fig:phylo_backbones}).
Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,431 sites.
% TODO this is out of date and needs to be rewritten along with supplementary 
% analysis
Remarkably, given the differences in inference methods and data processing pipelines,
they are almost identically parsimonious,
with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts).
We also compared the mutational spectra of major VOCs in the ARG
against the mutational spectra computed from the UShER phylogeny (STAR Methods),
finding a close match to previous reports\cite{Bloom2023}
(Figure~\ref{fig:mutational_spectra}).

@@

To evaluate the concordance of the methods in 
identifying the Pango lineages of recombinant parents and locating recombination breakpoints, 
we focus on the Pango Xs descending from the Type I recombination events (Table \ref{tab:pango_x_lineages}), 
using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt, we took available detection results for these Pango Xs reported in the original study;
for CovRecomb, we summarized the results from the original study (STAR Methods);
For each Pango X, we determined whether the inferred parental lineages and breakpoint intervals 
are concordant with those suggested by the community (STAR Methods).
We found that the methods performed well overall, 
with no evidence for significant difference in the concordance rate among them (Table \ref{tab:x_method_concordance}),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.

@@

\begin{table}
\caption{
Concordance among methods in characterizing Pango X lineages. 
The Pango X lineages analyzed here are associated with the Type 1 recombination events (related to Table 1).
For RecombinHunt and CovRecomb, the concordance rates are calculated 
based on the Pango Xs for which results are available from the original studies.
Also, for RecombinHunt, there are two sets of detection results, 
one set obtained using mutational profiles computed from a global GISAID dataset and another from a Nextstrain dataset.

}
\label{tab:method_concordance}
\begin{tabular}{llll}
\hline
\textbf{Method} & \textbf{Concordant in parental lineages (\%)} & \textbf{Concordant in breakpoint interval (\%)} & \textbf{Distance (bases)} \\ [1ex]
\hline
Sc2ts & 14 (93.3) & 13 (86.7) & 540.5 (315, 766) \\ [1ex]
\hline
RH-GISAID & 13 (100.0) & 10 (76.9) & 764.0 (4, 1202) \\ [1ex]
\hline
RH-Nextstrain & 12 (85.7) & 11 (78.6) & 1160.0 (559, 2785) \\ [1ex]
\hline
CovRecomb & 7 (87.5) & 4 (50.0) & 1562.5 (217, 13208) \\ [1ex]
\hline
\end{tabular}
\end{table}

@@

@@

\subsection{Major deletions in the ARG}\label{sec:supplement_dels}
% Referenced by the main Results.
Li et al. \cite{Li2023} performed a large-scale analysis of indels 
using a global GISAID dataset (sampled up to August, 2022). 
We examined the major indels reported in Table S1 of that study 
which have a minimum frequency of 0.01 
(sequence count divided by 9,149,680 filtered sequences).
Most of these indels (all deletions) are encoded in the ARG (Table~\ref{tab:indels}) 
and have high frequencies in the ARG
(sequence count divided by the total number of samples), 
except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
During site remapping, the 9-base deletion at 11283 was split up into 
a 5-base deletion at 11283 (count of 339867) and 
a 4-base deletion at 11288 (count of 871782).
The 686 deletion occurs below our threshold for inclusion into the ARG.

% See https://github.com/jeromekelleher/sc2ts-paper/issues/472
\begin{table}
\caption{Major deletions in Li et al. (2023).}
\label{tab:indels}
\begin{center}
\begin{tabular}{rrcrrrr}
\toprule & \multicolumn{4}{c}{Li et al.} & \multicolumn{2}{c}{Sc2ts ARG} \\
\cmidrule(lr){1-5} \cmidrule(lr){6-7}
Start & Length & Region & Count & Frequency & Count & Frequency \\
\midrule
28362 & 9 & N & 3939164 & 0.4305 & 871774 & 0.3512 \\
28271 & 1 & Intergenic & 3853692 & 0.4212 & 1118363 & 0.4506 \\
21765 & 6 & S & 3179887 & 0.3475 & 339681 & 0.1368 \\
11288 & 9 & ORF1ab-nsp6 & 3040576 & 0.3323 & 290408 & 0.1170 \\
28248 & 6 & ORF8 & 3026186 & 0.3307 & 1114669 & 0.4491 \\
22029 & 6 & S & 3010665 & 0.3290 & 1112410 & 0.4482 \\
11283 & 9 & ORF1ab-nsp6 & 2011934 & 0.2199 & 40 & 0.0000 \\
6513 & 3 & ORF1ab-nsp3 & 1984683 & 0.2169 & 340765 & 0.1373 \\
21633 & 9 & S & 1941962 & 0.2122 & 532072 & 0.2144 \\
21987 & 9 & S & 1918094 & 0.2096 & N/A & N/A \\
22194 & 3 & S & 1833050 & 0.2003 & 312706 & 0.1260 \\
21991 & 3 & S & 912944 & 0.0998 & 288831 & 0.1164 \\
686 & 9 & ORF1ab-nsp1 & 115450 & 0.0126 & N/A & N/A \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

@@

\begin{figure}
    \centering
    %TODO fix this to make the figures smaller.
    \includegraphics[width=\linewidth]{mutational_spectra}
    \caption{All-site mutational spectra of major VOCs.
(A) Mutational spectra of the ARG.
For each major VOC, the relative proportion of each mutation type was calculated from
the mutations occurring above the nodes with a Scorpio label associated with the VOC
(e.g. ``Alpha-like'' encompasses ``B.1.1.7-like'' and ``B.1.1.7-like+E484K'').
Note that $\sim$98\% of all the mutations in the ARG occur above the nodes
which have a Scorpio label associated with Alpha, Delta, Omicron BA.1, BA.2, BA.4, or BA.5
(excluding the nodes with unassigned scorpio labels).
The Scorpio labels were assigned to the aligned sequences of the nodes in the ARG using pangolin.
(B) Mutational spectra reproduced from the mutation count data analyzed in Bloom et al. (2023).
The authors extracted mutation counts from a global UShER phylogeny
(https://github.com/jbloomlab/SARS2-mut-spectrum/).
\label{fig:mutational_spectra}
}
\end{figure}

\begin{figure} 
\centering
\includegraphics[width=0.5\textwidth]{LS_model_schematic}
\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
model, in which a focal sequence (bottom) is described as an
imperfect mosaic of the sequences in a reference panel.
Black crosses along the focal sequence show sequencing
errors or mutations.
In the standard formulation, at site $\ell$, the recombination probability is $r_\ell$,
the mutation probability is $\mu_\ell$ and $n$
denotes the size of the reference panel.
The Viterbi algorithm can be used to find a
``copying path'' through the reference panel for a given focal sequence that
maximises the likelihood under these parameters. Unseen states in the reference panel are shown as coloured lines enclosed by
the grey box. The black arrow describes the true path through the data which leads to the emitted
focal sequence below. Examples of transition and
emission probabilities along this trajectory are shown by the red and blue
arrows, respectively.
}
\end{figure}

>>>>

The ARG represents the full evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024}.
First, we validated the broad structure of the ARG against the UShER phylogeny\cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
Tanglegrams reveal that the ARG captures known phylogenetic relationships
both within sublineages and at the level of major lineages (including VOCs),
with good agreement between the phylogenetic bipartitions in the ARG and the UShER 
phylogeny (Figure~\ref{fig:phylo_backbones}).
Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,431 sites.
% TODO this is out of date and needs to be rewritten along with supplementary 
% analysis
Remarkably, given the differences in inference methods and data processing pipelines,
they are almost identically parsimonious,
with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts).
We also compared the mutational spectra of major VOCs in the ARG against the mutational spectra 
from the Viridian UShER phylogeny\cite{Hunt2024} and a previous study\cite{Bloom2023} (STAR Methods),
finding a close match to them (Figure~\ref{fig:mutational_spectra}).

@@

To evaluate the concordance of the methods in 
identifying the Pango lineages of recombinant parents and locating recombination breakpoints, 
we focus on the Pango Xs associated with the Type I recombination events (Table \ref{tab:pango_x_lineages}), 
using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt, we took detection results for these Pango Xs reported in the original study \cite{Alfonsi2024};
for CovRecomb, we summarized the results from the original study \cite{Li2024} (STAR Methods).
For each method, we determined the Pango Xs for which the inferred parental lineages and breakpoint intervals 
are concordant with those identified by the community (STAR Methods).
We found that the methods performed well overall, 
with comparable concordance rates (Table \ref{tab:x_method_concordance}),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.

@@

\begin{table}
\caption{
Concordance among methods in characterizing Pango X lineages. 
The Pango X lineages analyzed here are associated with the Type 1 recombination events (related to Table 1).
For RecombinHunt and CovRecomb, the concordance rates are calculated 
based on the Pango Xs for which results are available from the original studies.
Also, for RecombinHunt, two sets of detection results are available, 
one set obtained using mutational profiles computed from a global GISAID dataset and another from a Nextstrain dataset.}
\label{tab:method_concordance}
\begin{tabular}{llll}
\hline
\textbf{Method} & \textbf{Concordant in parental lineages (\%)} & \textbf{Concordant in breakpoint interval (\%)} & \textbf{Distance (bases)} \\ [1ex]
\hline
Sc2ts & 14 (93.3) & 13 (86.7) & 540.5 (315, 766) \\ [1ex]
\hline
RH-GISAID & 13 (100.0) & 10 (76.9) & 764.0 (4, 1202) \\ [1ex]
\hline
RH-Nextstrain & 12 (85.7) & 11 (78.6) & 1160.0 (559, 2785) \\ [1ex]
\hline
CovRecomb & 7 (87.5) & 4 (50.0) & 1562.5 (217, 13208) \\ [1ex]
\hline
\end{tabular}
\end{table}

@@

\subsubsection{Mutational spectra}
To validate the mutational information contained in the sc2ts ARG,
we compared the all-site mutational spectra of major VOCs in the ARG,
a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024}, and
a global UShER phylogeny analyzed in a previous study\cite{Bloom2023}.
Mutational spectra are relative proportions of each type of nucleotide mutation.

First, we calculated the mutational spectra of Alpha, Delta, Omicron BA.1, BA.2, BA.4, and BA.5
in the sc2ts ARG, considering all the mutations above the sample nodes and internal nodes
which have Scorpio labels associated with the VOCs (e.g., B.1.1.7-like and B.1.1.7-like+E484K for Alpha).
These Scorpio labels were assigned to the aligned sequences of the nodes in the ARG using pangolin.
Note that $\sim$98\% of all the mutations in the ARG occur
above the nodes which have a Scorpio label associated with Alpha, Delta, BA.1, BA.2, BA.4, or BA.5
(excluding the nodes with unassigned Scorpio labels).

Next, we calculated the mutational spectra of the same VOCs
using the mutational counts from an UShER phylogeny analyzed in Bloom et al.\cite{Bloom2023},
which was built using Nextclade and has samples collected globally up to November, 2022.
We used the following mapping between the WHO labels and the Nextstrain labels:
Alpha, 20I; Detal, 21I and 21J; BA.1, 21K; BA.2, 21L; BA.4, 22A, and BA.5, 22B.
The mutation count data are available at https://github.com/jbloomlab/SARS2-mut-spectrum/.

Lastly, we calculated the mutational spectra of the same VOCs
in the Viridian UShER phylogeny\cite{Hunt2024},
which was pruned down to the samples present in the sc2ts ARG.
We used the mutations above the samples based on their Scorpio labels,
which are available in the Viridian metadata provided by Hunt et al.\cite{Hunt2024}.

\subsubsection{Concordance among recombination detection methods}
For RecombinHunt, we obtained two sets of detection results. 
One set is based on mutational profiles calculated from a global GISAID dataset, 
and the other set from a global Nextstrain dataset. 
We adjusted the RecombinHunt-identified breakpoint interval 
so that they are right-exclusive by adding one to the right coordinate.

For CovRecomb, we summarized the detection results for the Pango Xs as follows. 
First, we obtained the independent recombination events 
in Table S4 corresponding to a Pango X, if available. 
When multiple events corresponded to a Pango X, 
we took the results of the event having the most epidemic recombinants. 
Then, for other Pango Xs, we collated the data from a file (URL) 
containing all the detection results produced by the study. 
For each of these Pango Xs, 
we took the most commonly observed combination of 
left parent, right parent, and breakpoint interval.
We adjusted the CovRecomb-identified breakpoint interval 
so that they are right-exclusive by adding one to the right coordinate.

We ran rebar on the sampled sequences associated with each robust recombination event. 
Rebar identifies a sequence as a recombinant if it contains sets of mutations 
predetermined to be associated with different Pango lineages.

The parents of a Pango X inferred using a method are considered concordant 
if their Pango lineage assignments are at least as specific as the community-identified parents.
Also, a breakpoint interval of a Pango X inferred using a method is considered concordant 
if it overlaps with the community-identified breakpoint interval by at least one base.
We adjusted the breakpoint intervals from the community 
by subtract one from the left coordinate and adding two to the right coordinate.

@@

\subsection{Major deletions in the ARG}\label{sec:supplement_dels}
% Referenced by the main Results.
Li et al. \cite{Li2023} performed a large-scale analysis of indels 
using a global GISAID dataset (sampled up to August, 2022). 
We examined the major indels reported in Table S1 of that study 
which have a minimum frequency of 1\% 
(sequence count divided by 9,149,680 filtered sequences).
Most of these indels (all deletions) are encoded in the ARG (Table~\ref{tab:indels}) 
and have high frequencies in the ARG
(sequence count divided by the total number of samples), 
except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
During site remapping, the 9-base deletion at 11283 was split up into 
a 5-base deletion at 11283 (count of 339867) and 
a 4-base deletion at 11288 (count of 871782).
The 686 deletion occurs below our threshold for inclusion into the ARG.

% See https://github.com/jeromekelleher/sc2ts-paper/issues/472
\begin{table}
\caption{Major deletions in Li et al. (2023).}
\label{tab:indels}
\begin{tabular}{rrcrrrr}
\toprule & \multicolumn{4}{c}{Li et al. (2023)} & \multicolumn{2}{c}{Sc2ts ARG} \\
\cmidrule(lr){1-5} \cmidrule(lr){6-7}
Start & Length & Region & Count & Frequency (%) & Count & Frequency (%)\\
\midrule
28362 & 9 & N & 3939164 & 43.0525 & 871774 & 35.1216 \\
28271 & 1 & Intergenic & 3853692 & 42.1183 & 1118363 & 45.0561 \\
21765 & 6 & S & 3179887 & 34.7541 & 339681 & 13.6849 \\
11288 & 9 & ORF1ab-nsp6 & 3040576 & 33.2315 & 290408 & 11.6998 \\
28248 & 6 & ORF8 & 3026186 & 33.0742 & 1114669 & 44.9073 \\
22029 & 6 & S & 3010665 & 32.9046 & 1112410 & 44.8163 \\
11283 & 9 & ORF1ab-nsp6 & 2011934 & 21.9891 & 40 & 0.0016 \\
6513 & 3 & ORF1ab-nsp3 & 1984683 & 21.6913 & 340765 & 13.7286 \\
21633 & 9 & S & 1941962 & 21.2244 & 532072 & 21.4359 \\
21987 & 9 & S & 1918094 & 20.9635 & N/A & N/A \\
22194 & 3 & S & 1833050 & 20.0340 & 312706 & 12.5982 \\
21991 & 3 & S & 912944 & 9.9779 & 288831 & 11.6363 \\
686 & 9 & ORF1ab-nsp1 & 115450 & 1.2618 & N/A & N/A \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
    \centering
    %TODO fix this to make the figures smaller.
    \includegraphics[width=\linewidth]{mutational_spectra}
    \caption{All-site mutational spectra of major VOCs calculated from the sc2ts ARG (A),
    the mutation count data from Bloom et al. (2023) (B), and
    the Viridian UShER phylogeny from Hunt et al. (2024)\cite{Hunt2024}.
    \label{fig:mutational_spectra}
}
\end{figure}

\begin{figure} 
\centering
\includegraphics[width=0.5\textwidth]{LS_model_schematic}
\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
model, in which a focal sequence (bottom) is described as an
imperfect mosaic of the sequences in a reference panel.
Black crosses along the focal sequence show sequencing
errors or mutations.
In the standard formulation, at site $\ell$, the recombination probability is $r_\ell$,
the mutation probability is $\mu_\ell$ and $n$
denotes the size of the reference panel.
The Viterbi algorithm can be used to find a
``copying path'' through the reference panel for a given focal sequence that
maximises the likelihood under these parameters. Unseen states in the reference panel are shown as colored lines enclosed by
the grey box. The black arrow describes the true path through the data which leads to the emitted
focal sequence below. Examples of transition and
emission probabilities along this trajectory are shown by the red and blue
arrows, respectively.
}
\end{figure}
